---
title: "Activity Data Analysis"
author: "Timothy M. Rodriguez"
date: "Sunday, February 08, 2015"
output: html_document
---

##Getting the data
Here we download the data (if not already downloaded) and unzip it and report an overall summary of the data.

```{r}
file.name <- "activity"
file.name.zip <- paste(file.name, ".zip", sep="")
file.name.csv <- paste(file.name, ".csv", sep="")
if (!file.exists(file.name.zip)) {
    download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", file.name.zip)
}
unzip(file.name.zip)
activity.data <- read.csv(file.name.csv)

summary(activity.data)
```

##Answering Questions
Next, we'd like to answer some questions about the data.  For example

### 1. What is the mean total number of steps taken per day?

```{r}
hist(aggregate(steps ~ date, data=activity.data, sum)$steps,
     breaks=10,
     main="Histogram of steps taken per day",
     xlab="Steps taken")
median <- median(activity.data$steps, na.rm=T)
mean <- mean(activity.data$steps, na.rm=T)
```

The median was `r median` and the mean was `r mean`.

### 2. What is the average daily activity pattern?
Let's aggregate the data by interval.

```{r}
steps.by.interval <- aggregate(steps ~ interval, data=activity.data, sum)
plot(steps.by.interval$interval,
     steps.by.interval$steps,
     type="l",
     main="Average steps per day",
     xlab="Interval",
     ylab="Average Steps taken")
```

There may be some bias being introduced by missing values.

```{r}
number.of.missing.rows <- nrow(activity.data[!complete.cases(activity.data), ])

```
The total number of missing values are `r number.of.missing.rows`.

Let's try to create a data set with NAs replaced by the mean for that interval.

```{r}
get.mean.for.interval <- function(interval) {
    steps.by.interval[steps.by.interval$interval == interval, ]$steps
}
activity.data.no.nas <- data.frame(activity.data)
activity.data.no.nas$steps <- sapply(activity.data.no.nas$interval, get.mean.for.interval, simplify=T)
number.of.missing.rows.after.cleanup <- nrow(activity.data.no.nas[!complete.cases(activity.data.no.nas), ])
```

There are now `r number.of.missing.rows.after.cleanup` NAs after imputing values. Okay, let's plot the cleaned up data.

```{r}
steps.by.interval.no.nas <- aggregate(steps ~ interval, data=activity.data.no.nas, sum)
plot(steps.by.interval.no.nas$interval,
     steps.by.interval.no.nas$steps,
     type="l",
     main="Average steps per day",
     xlab="Interval",
     ylab="Average Steps taken")
```

That didn't seem to change the plot very much.

###3. Are there differences in activity patterns between weekdays and weekends?

Let's label each observation with a factor for weekend or weekday.

```{r}
day.type <- factor(c("weekday", "weekend"))
get.day.type <- function(date) {
    day <- weekdays(as.Date(date))
    if (day == "Saturday" || day == "Sunday") {
        levels(day.type)[2]
    } else {
        levels(day.type)[1]
    }
}
activity.data.no.nas$day.type <- sapply(activity.data.no.nas$date, get.day.type, simplify=T)
```

Next, let's aggregate the data by interval and day type.

```{r}
steps.by.interval.and.day.type <- aggregate(steps ~ interval + day.type, data=activity.data.no.nas, sum)
library(ggplot2)
ggplot(steps.by.interval.and.day.type, aes(x=interval, y=steps, group="weekday")) +
    ggtitle("Steps Taken on Weekdays") +
    geom_line() +
    facet_wrap(~day.type)
```

Here, we can see that activity drops markedly during the weekend.
